Why is it that statistical tests for residuals are not widely used? An explanation using visual inference.

### 2023-01-09 paper_comparison/paper_comparison/paper_comparison.tex
major update

Diagnosing a model is the key to determining whether there is anything
importantly wrong. For linear regression analysis, we typically
interrogate the residuals for model diagnostics. Residuals summarise
what is not captured by the model, and thus provide the capacity to
identify what might be wrong.

We can assess residuals in multiple ways. Residuals might be plotted, as
a histogram or quantile-quantile plot to examine the distribution. Using
the classical normal linear regression model as an example, if the
distribution is symmetric and unimodal, it is well-behaved. But if the
distribution is skewed, bimodal, multimodal, or contains outliers, there
is cause for concern. The distribution could also be inspected by
conducting a goodness of fit test, such as the Shapiro-Wilk Normality
test \citep{shapiro1965analysis}.

More typically, residuals will be plotted, as a scatter plot against the
predicted values and each of the explanatory variables to scrutinize
their relationships. If there is any visually discoverable patterns, the
model is potentially misspecified. However, correctly judging a residual
plot where no pattern exists is a painstakingly difficult task for
humans. It is especially common, particularly among new data analytics
students to report patterns when an experienced data analyst might
quickly conclude that there are none. Generally, one looks for
noticeable departures from the model like non-linear dependency or
heteroskedasticity. It is also possible to conduct hypothesis tests for
non-linear dependence \citep{ramsey_tests_1969}, and use a Breusch-Pagan
test \citep{breusch_simple_1979} for heteroskedasticity.

Abundance of literature describe appropriate diagnostic methods for
linear regression: \citet{draper1998applied},
\citet{montgomery1982introduction}, \citet{belsley_regression_1980},
\citet{cook_applied_1999} and \citet{cook1982residuals}. The diligent
reader of these sage writings will also notice sentences that express
sentiments like \emph{based on their experience, statistical tests are
not widely used in regression diagnostics. The same or even larger
amount of information can be provided by diagnostic plots than the
corresponding tests in most empirical studies.} A common guidance by
experts is that optimal method for diagnosing model fits is by plotting
the data.

The persistence of this advice to check the plots is curious, and
investigating why this might be common advice is the subject of this
paper. The paper is structured as follows. The next background section
describes the types of departures that one expects to detect, and
describes a formal statistical process for reading residual plots,
called visual inference. Section \ref{experiment-design} describes the
experiment design to compare the decision made by formal hypothesis
testing, and how humans would read diagnostic plots. The results are
reported in Section \ref{results}. We finish with a discussion on future
work, in particular how the responsibility for residual plot reading
might be passed on to computer vision.

@@

Graphical summaries in which residuals are plotted against fitted values
or other functions of the regressors that are approximately orthogonal
to residuals are referred to as standard residual plots in
\citet{cook1982residuals}. As shown in Figure
\ref{fig:residual-plot-common-departures}, the top-left panel is a good
residual plot with residuals evenly distributed at both sides of the
horizontal zero line showing no noticeable patterns.

There are various types of departures from a good residual plot.
Non-linearity, heterskedasticity and non-normality are perhaps the three
mostly checked departures.

@@

Other than checking diagnostic plots, analysts may perform formal
hypothesis testing for detecting model defects. Depending on the
alternative hypothesis that is focused on, a variety of tests can be
applied. For example, the presence of heteroskedasticity can usually be
tested by applying the White test
\citep{white_heteroskedasticity-consistent_1980} or the Breusch-Pagan
test \citep{breusch_simple_1979}, which are both derived from the
Lagrange multiplier test \citep{silvey1959lagrangian} principle that
relies on the asymptotic properties of the null distribution. For
testing non-linearity, one may apply the F-test as a model structural
test to examine the significance of specific polynomial and non-linear
forms of the regressors, or the significance of proxy variables as in
the Ramsey Regression Equation Specification Error Test (RESET)
\citep{ramsey_tests_1969}. And for testing normality, the Shapiro--Wilk
test \citep{shapiro1965analysis} is perhaps the most widely used test
included by many of the statistical softwares. Another choice will be
the Jarque--Bera test \citep{jarque1980efficient} which directly checks
if the sample skewness and kurtosis match a normal distribution.

Example residual plots given in Figure
\ref{fig:residual-plot-common-departures} are examined by the
corresponding RESET test, Breusch-Pagan test and Shapiro-Wilk test as
shown in Table \ref{tab:example-residual-plot-table}. In the example,
both the Breusch-Pagan test and the Shapiro--Wilk test rejects the null
hypothesis \(H_0\) for departures that they do not intend to examine. As
discussed in \citet{cook1982residuals}, most residual-based tests for a
particular type of departure from model assumptions are sensitive to
other types of departures. It is likely \(H_0\) is correctly rejected
but for the wrong reason, which is known as the ``Type III error''.
Additionally, outliers will often incorrectly trigger the rejection of
\(H_0\) despite when majority of the residuals are well-behaved
\citep{cook_applied_1999}. Furthermore, with a sufficiently large sample
size, residual-based tests may reject \(H_0\) due to a slight departure
that is of little practical significance. These can be largely avoided
in diagnostic plots as experienced analysts can evaluate the
acceptability of assumptions flexibly, even in the presence of outliers
and slight departures.

@@

One may argue that reading diagnostic plots is to some extent subjective
and indecisive compared to those rigorous statistical procedures as it
relies on graphical perception - human's ability to interpret and decode
the information embedded in the graph \citep{cleveland_graphical_1984}.
It is particularly true that visual discovery suffers from its unsecured
and unconfirmed nature where the degree of the presence of the visual
features typically can not be measured quantitatively and objectively,
which may lead to over or under-interpretations of the data. One such
example is finding an over-interpretation of the separation between gene
groups in a two-dimensional projection from a linear discriminant
analysis when in fact there are no differences in the expression levels
between the gene groups and separation is not an uncommon occurrence
\citep{roy_chowdhury_using_2015}.

Visual inference was first introduced in a 1999 Joint Statistical
Meetings (JSM) talk with the title ``Inference for Data Visualization''
by \citet{buja_inference_1999} as an idea to address the issue of valid
inference for visual discoveries of data plots
\citep{gelman_exploratory_2004}. Later, \citet{buja_statistical_2009}
proposed the lineup protocol as a visual test inspired by the ``police
lineup'' or ``identity parade'' which is the act of asking the
eyewitness to identify criminal suspect from a group of irrelevant
people. The protocol consists of \(m\) randomly placed plots, where one
plot is the actual data plot, and the remaining \(m - 1\) plots have the
identical graphical production as the data plot except the data has been
replaced with data consistent with \(H_0\). Then, an observer who have
not seen the actual data plot will be asked to point out the most
different plot from the lineup. Under \(H_0\), it is expected that the
actual data plot would have no distinguishable difference with the null
plots, and the probability of the observer correctly picks the actual
data plot is \(1/m\). If we reject \(H_0\) as the observer correctly
picks the actual data plot, then the Type I error of this test is
\(1/m\).

@@

\caption{Visual testing is conducted using a lineup, as in the example here. The residual plot computed from the observed data (plot $2^2 + 2$, exhibiting non-linearity) is embedded among 19 null plots, where the residuals were simulated from a standard error model. Computing the $p$-value requires that the lineup be examined by a number of human judges, each asked to select the most different plot. A small $p$-value would result from a substantial number selecting plot $2^2 + 2$.}\label{fig:first-example-lineup}
\end{figure}

Figure \ref{fig:first-example-lineup} is an example of a lineup
protocol. If the actual data plot at position \(2^2 + 2\) is
identifiable, then it is evidence for the rejection of \(H_0\) that the
regression model is correctly specified. In fact, the actual residual
plot is obtained from a misspecified regression model with non-linearity
defect.

The effectiveness of lineup protocol is validated by
\citet{majumder_validation_2013} under relatively simple classical
normal linear regression model settings with one or two regressors.
Their results suggest that visual test is capable of testing the
significance of a single regressor with a similar power as a t-test,
though they express that in general it is unnecessary to use visual
inference if there exists a conventional test and they do not expect the
visual test to perform equally well as the conventional test. In their
third experiment, where there does not exist a proper conventional test,
visual test outperforms the conventional test for a large margin. This
is encouraging as it promotes the use of visual inference in border
field of data science where there are no existing statistical testing
procedures. Visual inference have also already been integrated into
diagnostic of hierarchical linear models by \citet{loy2013diagnostic},
\citet{loy2014hlmdiag} and \citet{loy2015you}. They use lineup protocols
to judge the assumption of linearity, normality and constant error
variance for both the level-1 and level-2 residuals.

@@

Data used in the \(m - 1\) null plots need to be simulated. In the
context of regression diagnostics, sampling data from \(H_0\) is
equivalent to sampling data from the assumed model. As
\citet{buja_statistical_2009} suggested, \(H_0\) is usually composited
by a collection of distributions controlled by nuisance parameters.
Since regression models can have various forms, there is no general
solution to this problem, but it sometimes can be reduced to so called
``reference distribution'' by applying one of the three methods: (i)
sampling from a conditional distribution given a minimal sufficient
statistic under \(H_0\), (ii) parametric bootstrap sampling with
nuisance parameters estimated under \(H_0\), and (iii) Bayesian
posterior predictive sampling. The conditional distribution given a
minimal sufficient statistic is the best justified reference
distribution among the three \citep{buja_statistical_2009}. Essentially,
null residuals can be simulated by regressing \(N\) i.i.d standard
normal random draws on the regressors, then rescaling it by the ratio of
residual sum of square in two regressions.

@@

Further, a visual test can involve \(K\) independent observers. Let
\(X_i = \{0,1\}\) be a binomial random variable denoting whether subject
\(i\) correctly detecting the actual data plot, and
\(X = \sum_{i=1}^{K}X_i\) be the number of observers correctly picking
the actual data plot. Then, by imposing a relatively strong assumption
on the visual test that all \(K\) evaluations are fully independent,
under \(H_0\), \(X \sim \mathrm{Binom}_{K,1/m}\). Therefore, the
\(p\)-value of a lineup of size \(m\) evaluated by \(K\) observer is
given as \(P(X \geq x) = 1 - F(x) + f(x)\), where \(F(.)\) is the
cumulative distribution function, \(f(.)\) is the probability mass
function and \(x\) is the realization of number of observers correctly
picking the actual data plot \citep{majumder_validation_2013}.

As pointed out by \citet{vanderplas2021statistical}, this basic binomial
model doesn't take into account the possible dependencies in the visual
test due to repeated evaluations of the same lineup. And it is
inapplicable to visual test where subjects are asked to select one or
more ``most different'' plots from the lineup.
\citet{vanderplas2021statistical} summarises three common scenarios in
visual inference: (1) \(K\) different lineups are shown to \(K\)
subjects, (2) \(K\) lineups with different null plots but the same
actual data plot are shown to \(K\) subjects, and (3) the same lineup is
shown to \(K\) subjects. Out of these three scenarios, Scenario 3 is the
most common in previous studies as it puts the least constraints on the
experiment design. For Scenario 3, \citet{vanderplas2021statistical}
model the probability of a plot \(i\) being selected from a lineup as
\(\theta_i\), where \(\theta_i \sim Dirichlet(\alpha)\) for
\(i=1,...,m\) and \(\alpha > 0\). And defined \(c_i\) to be the number
of times plot \(i\) being selected in \(K\) evaluations. In case subject
\(j\) makes multiple selections, \(1/s_j\) will be added to \(c_i\)
instead of one, where \(s_j\) is the number of plots subject \(j\)
selected for \(j=1,...K\). This ensured \(\sum_{i}c_i=K\). Since we are
only interested in the selections of the actual data plot \(i\), the
marginal model can be simplified to a beta-binomial model and thus the
visual p-value is given as

@@

Note that Equation \ref{eq:pvalue-beta-binomial} only works with
non-negative integer \(c_i\). For non-negative real number \(c_i\),
linear approximation will be applied to calculate the p-value

@@

\begin{equation}
P(C = c_i) = {K \choose c_i}\frac{B(c_i + \alpha, K - c_i + (m - 1)\alpha)}{B(\alpha, (m-1)\alpha)},\quad \text{for} \quad c_i \in \mathbb{Z}_0^+.
\end{equation}

Besides, the parameter \(\alpha\) used in Equation
\ref{eq:pvalue-beta-binomial} is usually unknown and hence needs to be
estimated from the survey data. For low values of \(\alpha\), only a few
plots are attractive to the observers and tend to be selected. For
higher values of \(\alpha\), the distribution of the probability of each
plot being selected is more even. \citet{vanderplas2021statistical}
define that a plot is \(c\)-interesting if \(c\) or more participants
select the plot as the most different. Given the definition, The
expected number of plots selected at least \(c\) times, \(E[Z_c]\), is
calculated as

@@

\citet{vanderplas2021statistical} suggest that \(\alpha\) can be
estimated using MLE with Equation \ref{eq:c-interesting-expectation}.
But for precise estimate of \(\alpha\), additional responses to
Rorschach lineups, which is a type of lineup that consists of plots
constructed from the same null data generating mechanism, are required.

@@

As discussed in \citet{majumder_validation_2013}, individual's skill
will affect the number of observers who identify the actual data plot
from the lineup. Thus, the power of a visual test depends on the
subject-specific abilities. Previously, it is addressed by modelling the
probability of a subject \(i\) correctly picking the actual data plot
from a lineup \(l\) using a mixed-effect logistic regression with the
subject being treated as a random effect
\citep{majumder_validation_2013}. However, having this probability is
insufficient to determine the power of a visual test allowed for
multiple selections as it doesn't provide information about the number
of selections made by the subject for p-value calculation.

Instead, we omitted the individual effect and directly estimated the
probability of a lineup being rejected using a logistic regression with
the natural logarithm of the effect size as the only regressor
formulated as

\begin{equation} \label{eq:logistic-regression-1-1}
Pr(\text{reject}~H_0|H_1,E) = \Lambda(\beta_0 + \beta_1 log_e(\boldsymbol{E})),
\end{equation}

@@

Three experiments were conducted to investigate the difference between
conventional hypothesis testing and visual inference in the application
of linear regression diagnostics. The experiment I has ideal scenario
for conventional testing, where the visual test is not expected to
outperform the conventional test. Meanwhile, the experiment II is a
scenario where the conventional test is an approximate test, in which
the visual test may have a chance to match the performance of the
conventional test. The experiment III is designed for collecting human
responses to lineup with only good residual plots such that the
parameter \(\alpha\) in Equation \ref{eq:pvalue-beta-binomial} can be
estimated. Overall, we planned to collect 7974 evaluations on 1152
uniqued lineups performed by 443 subjects throughout three experiment.

@@

\begin{align} \label{eq:nonlinearity-model}
\boldsymbol{y} = 1 + \boldsymbol{x} + \boldsymbol{z} + \boldsymbol{\varepsilon},\\
\boldsymbol{x} = g(\boldsymbol{x}_{raw}, 1), \\
\boldsymbol{z} = g(\boldsymbol{z}_{raw}, 1), \\
\boldsymbol{z}_{raw} = He_j(g(\boldsymbol{z}, 2)),
\end{align}

\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}\),
\(\boldsymbol{\varepsilon}\), \(\boldsymbol{x}_{raw}\),
\(\boldsymbol{z}_{raw}\) are vectors of size \(n\), \(He_{j}(.)\) is the
\(j\)th-order probabilist's Hermite polynomials,
\(\varepsilon \sim N(\boldsymbol{0}, \sigma^2\boldsymbol{I}_n)\), and
\(g(\boldsymbol{x}, k)\) is a scaling function to enforce the support of
the random vector to be \(\{-k, k\}\) defined as

@@

Hermite polynomials were initially defined by \citet{de1820theorie}, but
named after Hermite \citep{hermite1864nouveau} because of the
unrecognisable form of Laplace's work. When simulating \(Z_{raw}\), the
function \texttt{hermite} from the R package \texttt{mpoly}
\citep{mpoly} was used to generate Hermite polynomials.

@@

Since \(z = O(x^j)\), for \(j > 1\), \(z\) is a higher order term leaves
out by the null regression, which will result in model misspecification.
Visual patterns of non-linearity were simulated using four different
order of probabilist's Hermite polynomials (\(j = 2, 3, 6, 18\)) and
four different distribution of \(X_{raw}\): (1) \(U(-1, 1)\), (2)
\(N(0, 0.3^2)\), (3) \(lognormal(0, 0.6^2)/3\) and (4) \(u\{1, 5\}\). A
summary of the factors is given in Table \ref{tab:model-factor-table}.

@@

The values of \(j\) was chosen so that distinct shapes of non-linearity
were included in the residual plot. A greater value of \(j\) will result
in a curve with more turning points. As shown in Figure
\ref{fig:different-shape-of-herimite}, it includes ``U'' shape, ``S''
shape, ``M'' shape and ``Triple-U'' shape. It is expected that the ``U''
shape will be the easiest one to detect because complex shape tends to
be concealed by cluster of data points.

Four different distribution were used to generate \(X_{raw}\) as shown
in Figure \ref{fig:different-dist}. The uniform and the normal
distribution are symmetric and commonly assumed in statistical models.
The adjusted log-normal distribution provides skewed density, while the
discrete uniform distribution provides discreteness in residual plot,
which could enrich the pool of visual patterns.

@@

Figure \ref{fig:example-poly-lineup} shows one of the lineups used in
experiment I. This lineup is produced under \(j = 6\) and
\(X_{raw} \sim N(0.0.3^2)\). The actual data plot location is
\(2^3 - 4\). All five subjects correctly identified the actual data plot
for this lineup.

@@

\begin{align} \label{eq:heter-model}
\boldsymbol{y} = 1 + \boldsymbol{x} + \boldsymbol{\varepsilon},\\
\boldsymbol{x} = g(\boldsymbol{x}_{raw}, 1)\\
\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}, 1 + 2 - |a| b (\boldsymbol{x} - a)^2 \boldsymbol{I}), \\
\end{align}

>>>>

Diagnosing a model is the key to determining whether there is anything
importantly wrong. In linear regression analysis, residuals are
typically examined for model diagnostics. Residuals summarise what is
not captured by the model, and thus provide the capacity to identify
what might be wrong.

One can access residuals in multiple ways. Residuals might be plotted,
as a histogram or quantile-quantile plot to examine the distribution.
Using the classical normal linear regression model as an example, if the
distribution is symmetric and unimodal, it is well-behaved. But if the
distribution is skewed, bimodal, multimodal, or contains outliers, there
is cause for concern. One could also inspect the distribution by
conducting a goodness of fit test, such as the Shapiro-Wilk Normality
test \citep{shapiro1965analysis}.

More typically, residuals will be plotted, as a scatter plot against the
predicted values and each of the explanatory variables to scrutinize
their relationships. If there are any visually discoverable patterns,
the model is potentially misspecified. In general, one looks for
noticeable departures from the model like non-linear dependency or
heteroskedasticity. However, correctly judging a residual plot where no
pattern exists is a painstakingly difficult task for humans (?citation).
It is especially common, particularly among new data analysts, to report
patterns when an experienced one might quickly conclude that there are
none. It is also possible to conduct hypothesis tests for non-linear
dependence \citep{ramsey_tests_1969}, and use a Breusch-Pagan test
\citep{breusch_simple_1979} for heteroskedasticity.

Abundance of literature describe appropriate diagnostic methods for
linear regression: \citet{draper1998applied},
\citet{montgomery1982introduction}, \citet{belsley_regression_1980},
\citet{cook_applied_1999} and \citet{cook1982residuals}. The diligent
reader of these sage writings will also notice sentences that express
sentiments like \emph{based on their experience, statistical tests are
not widely used in regression diagnostics since the same or even larger
amount of information can be provided by diagnostic plots than the
corresponding tests in most empirical studies.} A common guidance by
experts is that optimal method for diagnosing model fits is by plotting
the data.

The persistence of this advice to check the plots is curious, and
investigating why this might be common advice is the subject of this
paper. The paper is structured as follows. The next background section
describes the types of departures that one expects to detect, and
outlines a formal statistical process for reading residual plots, called
visual inference. Section \ref{experimental-design} details the
experimental design to compare the decision made by formal hypothesis
testing, and how humans would read diagnostic plots. The results are
reported in Section \ref{results}. We finish with a discussion on future
work, in particular how the responsibility for residual plot reading
might be relieved.

@@

Graphical summaries in which residuals are plotted against fitted values
or other functions of the regressors that are approximately orthogonal
to residuals are referred to as standard residual plots in
\citet{cook1982residuals}. As shown in Figure
\ref{fig:residual-plot-common-departures}, the plot A is an ideal
residual plot with residuals evenly distributed at both sides of the
horizontal zero line, with no noticeable patterns.

There are various types of departures from an ideal residual plot.
Non-linearity, heterskedasticity and non-normality are perhaps the three
mostly checked departures.

@@

Other than checking diagnostic plots, analysts may perform formal
hypothesis testing for detecting model defects. Depending on the
alternative hypothesis that is focused on, a variety of tests can be
applied. For example, the presence of heteroskedasticity can usually be
tested by applying the White test
\citep{white_heteroskedasticity-consistent_1980} or the Breusch-Pagan
test \citep{breusch_simple_1979}, which are both derived from the
Lagrange multiplier test \citep{silvey1959lagrangian} principle that
relies on the asymptotic properties of the null distribution. For
testing non-linearity, one may apply the F-test as a model structural
test to examine the significance of specific polynomial and non-linear
forms of the regressors, or the significance of proxy variables as in
the Ramsey Regression Equation Specification Error Test (RESET)
\citep{ramsey_tests_1969}. The Shapiro-Wilk test
\citep{shapiro1965analysis} is the most widely used test of
non-normality included by many of the statistical softwares. The
Jarque--Bera test \citep{jarque1980efficient} is also used to directly
checks if the sample skewness and kurtosis match a normal distribution.

Example residual plots given in Figure
\ref{fig:residual-plot-common-departures} are examined by the
corresponding RESET test, Breusch-Pagan test and Shapiro-Wilk test as
shown in Table \ref{tab:example-residual-plot-table}. In the example,
the Breusch-Pagan test and the Shapiro--Wilk test both reject the null
hypothesis \(H_0\) for departures that they do not intend to examine. As
discussed in \citet{cook1982residuals}, most residual-based tests for a
particular type of departure from model assumptions are also sensitive
to other types of departures. It is likely \(H_0\) is correctly rejected
but for the wrong reason, a phenomenon known as the ``Type III error''.
Additionally, outliers will often incorrectly trigger the rejection of
\(H_0\) despite when majority of the residuals are well-behaved
\citep{cook_applied_1999}. Furthermore, with a sufficiently large sample
size, residual-based tests may reject \(H_0\) due to a slight departure
that is of little practical significance. These can be largely avoided
in diagnostic plots as experienced analysts can evaluate the
acceptability of assumptions flexibly, even in the presence of outliers
and slight departures.

@@

One may argue that reading diagnostic plots is to some extent subjective
and indecisive compared to those rigorous statistical procedures as it
relies on graphical perception - human ability to interpret and decode
the information embedded in graph \citep{cleveland_graphical_1984}.
Further, the degree of the presence of the visual features typically can
not be measured quantitatively and objectively, which may lead to over
or under-interpretations of the data. For instance, people
over-interpret the separation between gene groups in a two-dimensional
projection from a linear discriminant analysis when in fact there are no
differences in the expression levels between the gene groups and
separation is not an uncommon occurrence
\citep{roy_chowdhury_using_2015}.

Visual inference was first introduced in a 1999 Joint Statistical
Meetings (JSM) talk with the title ``Inference for Data Visualization''
by \citet{buja_inference_1999} as an idea to address the issue of valid
inference for visual discoveries of data plots. Later,
\citet{buja_statistical_2009} proposed the lineup protocol as a visual
test inspired by the ``police lineup'' or ``identity parade'' which is
the act of asking the eyewitness to identify criminal suspect from a
group of irrelevant people. The protocol consists of \(m\) randomly
placed plots, where one plot is the data plot, and the remaining
\(m - 1\) plots have the identical graphical procedure except the data
has been replaced with data consistent with \(H_0\). Then, an observer
who have not seen the data plot will be asked to point out the most
different plot from the lineup. Under \(H_0\), it is expected that the
data plot would have no distinguishable difference from the null plots,
and the probability that the observer correctly picks the data plot is
\(1/m\). If one rejects \(H_0\) as the observer correctly picks the data
plot, then the Type I error of this test is \(1/m\).

@@

\caption{Visual testing is conducted using a lineup, as in the example here. The residual plot computed from the observed data (plot $2^2 + 2$, exhibiting non-linearity) is embedded among 19 null plots, where the residuals are simulated from a standard error model. Computing the $p$-value requires that the lineup be examined by a number of human judges, each asked to select the most different plot. A small $p$-value would result from a substantial number selecting plot $2^2 + 2$.}\label{fig:first-example-lineup}
\end{figure}

Figure \ref{fig:first-example-lineup} is an example of a lineup
protocol. If the data plot at position \(2^2 + 2\) is identifiable, then
it is evidence for the rejection of \(H_0\) that the regression model is
correctly specified. In fact, the actual residual plot is obtained from
a misspecified regression model with non-linearity defect.

The effectiveness of lineup protocol for regression analysis is
validated by \citet{majumder_validation_2013} under relatively simple
settings with up to two regressors. Their results suggest that visual
tests are capable of testing the significance of a single regressor with
a similar power as a t-test, though they express that in general it is
unnecessary to use visual inference if there exists a conventional test,
and they do not expect the visual test to perform equally well as the
conventional test. In their third experiment, where there is not a
conventional test, visual test outperforms the conventional test for a
large margin. This is encouraging, as it promotes the use of visual
inference in situations where there are no existing statistical testing
procedures. Visual inference have also been integrated into diagnostic
of hierarchical linear models by \citet{loy2013diagnostic},
\citet{loy2014hlmdiag} and \citet{loy2015you}. They use lineup protocols
to judge the assumption of linearity, normality and constant error
variance for both the level-1 and level-2 residuals. (expand?)

@@

Data used in the \(m - 1\) null plots needs to be simulated. In the
context of regression diagnostics, sampling data from \(H_0\) is
equivalent to sampling data from the assumed model. As
\citet{buja_statistical_2009} suggested, \(H_0\) is usually a composite
hypothesis controlled by nuisance parameters. Since regression models
can have various forms, there is no general solution to this problem,
but it sometimes can be reduced to a so called ``reference
distribution'' by applying one of the three methods: (i) sampling from a
conditional distribution given a minimal sufficient statistic under
\(H_0\), (ii) parametric bootstrap sampling with nuisance parameters
estimated under \(H_0\), and (iii) Bayesian posterior predictive
sampling. The conditional distribution given a minimal sufficient
statistic is the best justified reference distribution among the three
\citep{buja_statistical_2009}. Essentially, null residuals can be
simulated by regressing \(N\) i.i.d standard normal random draws on the
regressors, then rescaling it by the ratio of residual sum of square in
two regressions.

@@

In hypothesis testing, \(p\)-value is defined as the probability of
observing test results as least as extreme as the observed result given
\(H_0\) is true. Within the context of visual inference, by involving
\(k\) independent observers, \(p\)-value can be interpreted as the
probability of having as many or more subjects detect the data plot than
the observed result.

Let \(X_j = \{0,1\}\) be a binomial random variable denoting whether
subject \(j\) correctly detecting the data plot, and
\(X = \sum_{j=1}^{K}X_j\) be the number of observers correctly picking
the data plot. Then, by imposing a relatively strong assumption on the
visual test that all \(K\) evaluations are fully independent, under
\(H_0\), \(X \sim \mathrm{Binom}_{K,1/m}\). Therefore, the \(p\)-value
of a lineup of size \(m\) evaluated by \(K\) observer is given as
\(P(X \geq x) = 1 - F(x) + f(x)\), where \(F(.)\) is the cumulative
distribution function, \(f(.)\) is the probability mass function and
\(x\) is the realization of number of observers correctly picking the
data plot \citep{majumder_validation_2013}.

As pointed out by \citet{vanderplas2021statistical}, this basic binomial
model doesn't take into account the possible dependencies in the visual
test due to repeated evaluations of the same lineup. And it is
inapplicable to visual test where subjects are asked to select one or
more ``most different'' plots from the lineup.
\citet{vanderplas2021statistical} summarises three common scenarios in
visual inference: (1) \(K\) different lineups are shown to \(K\)
subjects, (2) \(K\) lineups with different null plots but the same data
plot are shown to \(K\) subjects, and (3) the same lineup is shown to
\(K\) subjects. Out of these three scenarios, Scenario 3 is the most
common in previous studies as it puts the least constraints on the
experiment design. For Scenario 3, \citet{vanderplas2021statistical}
models the probability of a plot \(i\) being selected from a lineup as
\(\theta_i\), where \(\theta_i \sim Dirichlet(\alpha)\) for
\(i=1,...,m\) and \(\alpha > 0\). The number of times plot \(i\) being
selected in \(K\) evaluations is denoted as \(c_i\). In case subject
\(j\) makes multiple selections, \(1/s_j\) will be added to \(c_i\)
instead of one, where \(s_j\) is the number of plots subject \(j\)
selected for \(j=1,...K\). This ensures \(\sum_{i}c_i=K\). Since we are
only interested in the selections of the data plot \(i\), the marginal
model can be simplified to a beta-binomial model and thus the visual
p-value is given as

@@

Note that Equation \ref{eq:pvalue-beta-binomial} given in
\citet{vanderplas2021statistical} only works with non-negative integer
\(c_i\). We extend the equation to non-negative real number \(c_i\) by
applying a linear approximation

@@

\begin{equation} \label{eq:pmf-beta-binomial}
P(C = c_i) = {K \choose c_i}\frac{B(c_i + \alpha, K - c_i + (m - 1)\alpha)}{B(\alpha, (m-1)\alpha)},\quad \text{for} \quad c_i \in \mathbb{Z}_0^+.
\end{equation}

Besides, the parameter \(\alpha\) used in Equation
\ref{eq:pvalue-beta-binomial} and \ref{eq:pmf-beta-binomial} is usually
unknown and hence needs to be estimated from the survey data. For low
values of \(\alpha\), only a few plots are attractive to the observers
and tend to be selected. For higher values of \(\alpha\), the
distribution of the probability of each plot being selected is more
even. \citet{vanderplas2021statistical} defines that a plot is
\(c\)-interesting if \(c\) or more participants select the plot as the
most different. Given the definition, The expected number of plots
selected at least \(c\) times, \(E[Z_c]\), is calculated as

@@

With Equation \ref{eq:c-interesting-expectation}, \(\alpha\) can be
estimated using maximum likelihood estimation. But for precise estimate
of \(\alpha\), additional responses to Rorschach lineups, which is a
type of lineup that consists of plots constructed from the same null
data generating mechanism, are required.

@@

The power of a model misspecification test specifies the chance of a
misspecified regression model being detected given a certain alternative
hypothesis. It is an important indicator when one concerns if the
regression model is misspecified. Though, people might usually be more
interested in knowing how much the residuals depart from the model
assumptions and if the departure is of practical significance.

As discussed in \citet{majumder_validation_2013}, individual's skill may
affect the number of observers who identify the data plot from the
lineup. Thus, the power of a visual test could depend on the
subject-specific abilities. Previously, it is addressed by modelling the
probability of a subject \(j\) correctly picking the data plot from a
lineup \(l\) using a mixed-effect logistic regression with the subject
being treated as a random effect \citep{majumder_validation_2013}.
However, in the multiple selections scenario, having this probability is
insufficient to determine the power of a visual test as it doesn't
provide information about the number of selections made by the subject
for p-value calculation.

Instead, we estimate the probability of a lineup being rejected directly
by assuming the individual skill has negligible impact on the variation
of the power of a lineup. The assumption is made to simplify the model
structure, thus mitigating the need for expensive large-scale
experiments to estimate the complex covariance matrix. The model is a
logistic regression with the natural logarithm of the effect as the only
regressor formulated as

\begin{equation} \label{eq:logistic-regression-1-1}
Pr(\text{reject}~H_0|H_1,E) = \Lambda(\beta_0 + \beta_1 log_e(E)),
\end{equation}

@@

Three experiments are conducted to investigate the difference between
conventional hypothesis testing and visual inference in the application
of linear regression diagnostics. The experiment I has ideal scenario
for conventional testing, where the visual test is not expected to
outperform the conventional test. Meanwhile, the experiment II is a
scenario where the conventional test is an approximate test, in which
the visual test may have a chance to match the performance of the
conventional test. The experiment III is designed for collecting human
responses to null lineups such that the parameter \(\alpha\) in Equation
\ref{eq:pvalue-beta-binomial} can be estimated. Overall, we plan to
collect 7974 evaluations on 1152 uniqued lineups performed by 443
subjects throughout three experiments.

@@

\begin{align} \label{eq:nonlinearity-model}
\boldsymbol{y} = 1 + \boldsymbol{x} + \boldsymbol{z} + \boldsymbol{\varepsilon},\\
\boldsymbol{x} = g(\boldsymbol{x}_{raw}, 1), \\
\boldsymbol{z} = g(\boldsymbol{z}_{raw}, 1), \\
\boldsymbol{z}_{raw} = He_j(g(\boldsymbol{x}, 2)),
\end{align}

\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}\),
\(\boldsymbol{\varepsilon}\), \(\boldsymbol{x}_{raw}\),
\(\boldsymbol{z}_{raw}\) are vectors of size \(n\), \(He_{j}(.)\) is the
\(j\)th-order probabilist's Hermite polynomials,
\(\varepsilon \sim N(\boldsymbol{0}, \sigma^2\boldsymbol{I}_n)\), and
\(g(\boldsymbol{x}, k)\) is a scaling function to enforce the support of
the random vector to be \([-k, k]^n\) defined as

@@

According to \citet{abramowitz1964handbook}, Hermite polynomials were
initially defined by \citet{de1820theorie}, but named after Hermite
\citep{hermite1864nouveau} because of the unrecognisable form of
Laplace's work. When simulating \(\boldsymbol{z}_{raw}\), function
\texttt{hermite} from the R package \texttt{mpoly} \citep{mpoly} is used
to generate Hermite polynomials.

@@

Since \(z = O(x^j)\), for \(j > 1\), \(z\) is a higher order term leaves
out by the null regression, which will lead to model misspecification.

Visual patterns of non-linearity are simulated using four different
orders of probabilist's Hermite polynomials (\(j = 2, 3, 6, 18\)) and
four different distributions of \(X_{raw}\): (1) \(U(-1, 1)\), (2)
\(N(0, 0.3^2)\), (3) \(lognormal(0, 0.6^2)/3\) and (4) \(u\{1, 5\}\). A
summary of the factors is given in Table \ref{tab:model-factor-table}.

@@

The values of \(j\) is chosen so that distinct shapes of non-linearity
are included in the residual plot. These include ``U'', ``S'', ``M'' and
``Triple-U'' shape as shown in Figure
\ref{fig:different-shape-of-herimite}. A greater value of \(j\) will
result in a curve with more turning points. It is expected that the
``U'' shape will be the easiest one to detect because complex shape
tends to be concealed by cluster of data points.

Different distributions of \(X_{raw}\) help enriching the pool of visual
patterns as illustrated in Figure \ref{fig:different-dist}. The uniform
and the normal distribution are symmetric and commonly assumed in
statistical models. The adjusted log-normal distribution provides skewed
density, while the discrete uniform distribution provides discreteness
in residual plot.

@@

Figure \ref{fig:example-poly-lineup} demonstrates one of the lineups
used in experiment I. This lineup is produced by the non-linearity model
under \(j = 6\) and \(X_{raw} \sim N(0.0.3^2)\). The data plot location
is \(2^3 - 4\). All five subjects correctly identify the data plot from
this lineup.

@@

\begin{align} \label{eq:heter-model}
\boldsymbol{y} = 1 + \boldsymbol{x} + \boldsymbol{\varepsilon},\\
\boldsymbol{x} = g(\boldsymbol{x}_{raw}, 1)\\
\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}, 1 + (2 - |a|)(\boldsymbol{x} - a)^2b \boldsymbol{I}), \\
\end{align}

\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}\),
\(\boldsymbol{\varepsilon}\) are vectors of size \(n\) and \(g(.)\) is
the scaling function defined in Equation \ref{eq:scaling-function}.
